WSClean: an implementation of a fast, generic wide-field imager

### 2014-03-11 articles/2014/Article-WSClean/wsclean.tex
Implemented Randall's comments

\section{Introduction}
Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.

@@

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, with the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

@@

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires a factor of two fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phase shift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer. We will refer to this operation as prediction.

@@

Using too few $w$-layers results in decorrelation of off-axis sources in the longer baselines. Additionally, $w$-aliasing errors can cause ghost sources in the image. An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}. When Cotton-Schwab cleaning includes prediction with too few $w$-layers, incorrect values will be subtracted from the visibilities even when no aliased sources are cleaned during minor iterations. Therefore, accurate prediction is more important than accurate imaging, because aliasing artefacts are attenuated by cleaning as long as the model is subtracted accurately.

@@

Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms. For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$. The time complexity for $w$-stacking is then given by
\begin{equation}
\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_{\max} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
\end{equation}
and for $w$-projection it is
\begin{equation}
\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_{\max}^2 \sin^2 \alpha_\textrm{FOV}\right).
\end{equation}
From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are the dominant expense. In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=13.5cm]{img/fullsky-together}
\caption{A beam-corrected MWA image of a 112~s zenith observation that covers almost the full sky. The lower image shows a zoom-in on the southern side lobe. PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved. The noise level in the side lobe is about 200 mJy. PKS J2358-6054 has been cleaned, but some artefacts remain because no direction-dependent calibration has been performed.}
\label{fig:full-sky-example}
\end{center}
\end{figure*}

@@

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a FITS file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as \textsc{kvis} \citep{karma-1996} and \textsc{ds9} support such FITS files and display their coordinates correctly. Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{wsclean}. The pulsar (centre of image) displays 17\% circular polarization. Due to inaccuracies in the current beam model, other (unpolarized) sources show $\sim$1\% leakage into Stokes~V.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beam-formed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all four combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, \textsc{wsclean} can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources. The pipeline includes \textsc{wsclean} to make the Stokes-parameter images.

@@

\subsection{Gridding} \label{sec:gridding}
A gridding convolution kernel improves the accuracy of gridding in $uv$-space. A common kernel function is a windowed sinc function, which acts as a low-pass filter. This decreases the flux of sources outside the FOV, and thus helps to attenuate aliased ghost sources and sidelobes \citep{post-correlation-filtering}. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

@@

In the $\textrm{ZA}=10\degree$ case, \textsc{casa} is slightly less accurate. The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}. An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed. Its accuracy is higher compared to the normal projection, but its residual noise is also higher. The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy. We have performed tests with and without the optimization of \S\ref{sec:decreasing-inversion-resolution}. They yield identical numbers.

@@

Panel~\ref{fig:timing-nsamples} shows the dependency on the size of the visibility set. Results for imaging at zenith and 10\degree ZA are shown. The size of the visibility set was varied by changing the time resolution, which affects the number of visibilities to be gridded without changing the maximum $w$-value. For larger sets, both methods show a linear time dependency on the number of visibilities. This implies that gridding or reading dominates the cost. In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.6 times faster at $\textrm{ZA}=0\degree$. With small data volumes, the FFTs start to dominate the cost, visible in Fig.~\ref{fig:timing-nsamples} as a flattening towards the left. At that point, \textsc{wsclean} is in both cases approximately three times faster. At zenith, the two methods are expected to perform almost identical. The factor of 2-3 difference in Fig.~\ref{fig:timing-nsamples} could be due to different choices in optimizations.

@@

In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the FOV. The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 0\degree ZA than $w$-projection. Imaging the MWA primary beam requires approximately an image size of 3k pixels for cleaning. If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1.5k. At ZA=10\degree, this saves about a factor of 2 in computing cost. The benefit of the optimization increases with resolution. At an image size of 10k pixels, cost is decreased by an order of magnitude.

@@

The results are summarized in Table~\ref{tbl:computational-cost-predictions}. \textsc{wsclean} is in most situations predicted to be 2--3 times faster than \textsc{casa}. \textsc{wsclean} has the largest benefit on the full LOFAR, MWA and the full SKA, where \textsc{wsclean} is respectively 7, 8 and 12 times faster. The $w$-snapshot method does not improve much over normal \textsc{wsclean} operation in any of the cases. The $w$-snapshot might become more valuable at higher ZAs. The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}. All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.

>>>>

\section{Introduction}
Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. The image degradation effects of the $w$-term are amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.

@@

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

@@

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Therefore, each pixel is divided by the weighted number of samples. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires a factor of two fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~multiply the image with the appropriate factor; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phase shift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer. We will refer to this operation as prediction.

@@

Using too few $w$-layers results in decorrelation of sources far from the phase centre in the longer baselines. Additionally, $w$-aliasing errors can cause ghost sources in the image. An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}. When Cotton-Schwab cleaning includes prediction with too few $w$-layers, incorrect values will be subtracted from the visibilities even when no aliased sources are cleaned during minor iterations. Therefore, accurate prediction is more important than accurate imaging, because aliasing artefacts are attenuated by cleaning as long as the model is subtracted accurately.

@@

Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms. For comparison, we can assume the antialising kernel can be neglected and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$. The time complexity for $w$-stacking is then given by
\begin{equation}
\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_{\max} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
\end{equation}
and for $w$-projection it is
\begin{equation}
\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_{\max}^2 \sin^2 \alpha_\textrm{FOV}\right).
\end{equation}
From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are the dominant expense. In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=13.5cm]{img/fullsky-together}
\caption{A beam-corrected MWA image of a 112~s zenith observation at 180~MHz that covers almost the full sky. The lower image shows a zoom-in on the southern side lobe. PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved. The noise level in the side lobe is about 200 mJy. PKS J2358-6054 has been cleaned, but some artefacts remain because no direction-dependent calibration has been performed.}
\label{fig:full-sky-example}
\end{center}
\end{figure*}

@@

A recentred image is in the projection of the zenith tangent plane and, like warped snapshots, will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a FITS file with the orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as \textsc{kvis} \citep{karma-1996} and \textsc{ds9} support such FITS files and display their coordinates correctly. Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{wsclean}. The pulsar (centre of image) displays 17\% circular polarization. Due to inaccuracies in the current beam model, other (unpolarized) sources show $\sim$1\% leakage into Stokes~V.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beam-formed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrix, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all four combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, \textsc{wsclean} can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources. The pipeline includes \textsc{wsclean} to make the Stokes-parameter images.

@@

\subsection{Gridding} \label{sec:gridding}
A gridding convolution kernel improves the accuracy of gridding in $uv$-space \citep{optimal-gridding-schwab-1983}. A common kernel function is a windowed sinc function, which acts as a low-pass filter. This decreases the flux of sources outside the FOV, and thus helps to attenuate aliased ghost sources and sidelobes \citep{post-correlation-filtering}. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

@@

In the $\textrm{ZA}=10\degree$ case, \textsc{casa} is slightly less accurate. The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}. An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed. As can be seen in Table~\ref{tbl:accuracy-measurements}, the source fluxes in the recentred image have smaller errors compared to the normal projection, but the residual noise is higher. The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy. We have performed tests with and without the optimization of \S\ref{sec:decreasing-inversion-resolution}. They yield identical numbers.

@@

Panel~\ref{fig:timing-nsamples} shows the dependency on the size of the visibility set. Results for imaging at zenith and 10\degree ZA are shown. The size of the visibility set was varied by changing the time resolution, which affects the number of visibilities to be gridded without changing the maximum $w$-value. For larger sets, both methods show a linear time dependency on the number of visibilities. This implies that gridding or reading dominates the cost. In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.6 times faster at $\textrm{ZA}=0\degree$. With small data volumes, the FFTs start to dominate the cost, visible in Fig.~\ref{fig:timing-nsamples} as a flattening towards the left. At that point, \textsc{wsclean} is in both cases approximately three times faster. At zenith, the two methods are expected to perform almost identically. The factor of 2-3 difference in Fig.~\ref{fig:timing-nsamples} could be due to different choices in optimizations.

@@

In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the FOV. The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 0\degree ZA than $w$-projection. Imaging the MWA primary beam requires approximately an image size of 3072 pixels for cleaning. If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1500. At ZA=10\degree, this saves about a factor of 2 in computing cost. The benefit of the optimization increases with resolution. At an image size of 10,000 pixels, cost is decreased by an order of magnitude.

@@

The results are summarized in Table~\ref{tbl:computational-cost-predictions}. \textsc{wsclean} is in most situations predicted to be 2--3 times faster than \textsc{casa}. \textsc{wsclean} has the largest benefit on the full LOFAR, MWA and the full SKA, where \textsc{wsclean} is respectively 7, 8 and 12 times faster. The $w$-snapshot method does not improve much over normal \textsc{wsclean} operation in any of the cases. The $w$-snapshot might become more valuable at higher ZAs. The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}. All configurations listed in Table~\ref{tbl:computational-cost-configurations}, with the exception of VLSS and GLEAM, have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
